---
title: "Verify the Claim: Rise of poverty rates in Philadelphia Neighbourhoods"
author: "Lauryn Chan"
date: 2026-9-14
format:
  html:
    toc: true
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
---

# Part 1 — Reconstruct

The WHYY article claims that poverty has risen from 2006-10 to 2013-17 based on the ACS five-year estimate. They have also quoted several neighbourhoods and calculated corresponding figures to prove their claim.

To verify their claims, I will be pulling the exact data that they used to reconstruct (hopefully) similar figures:

* **Years and survey**: ACS5 for 2010 and 2017
  * The two periods do not overlap as there is a gap between 2010 (the end of the first survey) and 2013 (the start of the second survey). This will avoid error generated from comparing the same information twice and does not display the full change in characteristic between the 2 years.
* **Variable code**: B07012_001, B07012_002
  * Based on the article that mentions "poverty rate" with margins of error, we need to compute poverty rate from the number of people below the poverty line (_002) divided by total population (_001).
* **Level**: Tract
  * To roughly equate tracts to neighbourhoods that were compared in the article

**Assumptions**: For simplistic reasons, I assume 1 tract will correspond to 1 neighbourhood referred to in the article. The corresponding tract number is found by inputting a random address in a residential area of the referenced neighbourhood into the [Census Geocoder](https://geocoding.geo.census.gov/geocoder/).


***

### 1a. Pulling data

```{r}
#| include: false
library(tidyverse)
library(tidycensus)
library(dplyr)
library(gt)
library(ggplot2)
```

```{r}
#| message: false
#| warning: false

my_state <- "PA"
my_county <- "Philadelphia"

get_poverty <- function(yr) {
  get_acs(
    geography = "tract",
    variables = c(below_poverty = "B07012_002", #this targets the people below the poverty line
                  total = "B07012_001"),
    state = my_state,
    county = my_county,
    year = yr,
    survey = "acs5",
    output = "wide"
  ) |> 
    mutate(period = paste0(yr - 4, "_", yr))
}

# Fetch ACS data
poverty_change <- bind_rows(
  get_poverty(2010),
  get_poverty(2017)
) |> 
  mutate(NAME = str_remove(NAME, ", Philadelphia County, Pennsylvania"))

poverty_change$period <- substr(poverty_change$period, 6, nchar(poverty_change$period))

head(poverty_change)

```
[_Above: Long format dataframe of ACS5 poverty data from 2006-2010 and 2013-2017_]{style="color: green;"}



***
# Part 2 — Test

### 2a. Computing the poverty rate

```{r}

# compute poverty share & moe before creating wide dataframe
poverty_change <- poverty_change |> 
  mutate(
    share     = below_povertyE / totalE,
    share_moe = moe_prop(below_povertyE, totalE,
                         below_povertyM, totalM)
  )

# wrangling to wide format to MASTER DATAFRAME
poverty_master <- poverty_change |>
  select(GEOID, NAME, period, share, share_moe) |> 
  rename(moe = share_moe) |> 
  pivot_wider(
    names_from = period,
    values_from = c(share, moe)
  )

head(poverty_master)

```
[_Above: Wide format dataframe of change in ACS5 poverty rate from 2006-2010 and 2013-2017_]{style="color: green;"}


### 2b. Coefficient of variation

```{r}

poverty_variation <- poverty_master |> 
  mutate(
    cv_2010 = (moe_2010 / 1.645)/share_2010 *100,
    cv_2017 = (moe_2017 / 1.645)/share_2017 *100,
  )

head(poverty_variation)

```
[_Above: Wide format dataframe of change in ACS5 poverty rate from 2006-2010 and 2013-2017_]{style="color: green;"}

### 2c. Reliability table

:::{style="max-height: 400px; overflow-y: auto;"}

```{r}


reliability_table <- poverty_variation |> 
  mutate(
    dot_2010 = case_when(
      cv_2010 < 12  ~ "🟢",
      cv_2010 <= 40 ~ "🟡",
      TRUE          ~ "🔴"
    ),
    dot_2017 = case_when(
      cv_2017 < 12  ~ "🟢",
      cv_2017 <= 40 ~ "🟡",
      TRUE          ~ "🔴"
    ),
    flag_2010 = case_when(
      cv_2010 < 12  ~ "Reliable",
      cv_2010 <= 40 ~ "Somewhat reliable",
      TRUE          ~ "Unreliable"
    ),
    flag_2017 = case_when(
      cv_2017 < 12  ~ "Reliable",
      cv_2017 <= 40 ~ "Somewhat reliable",
      TRUE          ~ "Unreliable"
    )
  )


# Make it look nice using gt package
reliability_table |> 
  gt() |> 
  fmt_percent(columns = c(share_2010, share_2017), decimals = 1) |> 
  fmt_number(columns = c(moe_2010, moe_2017), decimals = 3) |> 
  fmt_number(columns = c(cv_2010, cv_2017), decimals = 1) |> 
  cols_label(
    dot_2010 = "",
    dot_2017 = "",
    NAME     = "County",
    share_2010 = "Poverty Share 2010",
    share_2017 = "Poverty Share 2017",
    moe_2010 = "MOE 2010",
    moe_2017 = "MOE 2017",
    cv_2010  = "CV 2010 (%)",
    cv_2017  = "CV 2017 (%)",
    flag_2010 = "Reliability 2010",
    flag_2017 = "Reliability 2017"
  ) |> 
  cols_align(align = "center", columns = c(dot_2010, dot_2017)) |> 
  tab_header(
    title    = "Population below 100% of the poverty level",
    subtitle = "Philadelphia tracts, ACS 5-year estimates"
  ) |> 
  tab_source_note(
    "Reliability thresholds follow Jurjevich et al. (2018): CV < 12% reliable, 12–40% somewhat reliable, > 40% unreliable."
  )
```

:::
[_Above: Reliability table of poverty share per tract, based on coefficient of variation_]{style="color: green;"}


### 2d. Computing poverty change, and associated margin of error

I will test whether the change is statistically distinguishable from zero at 90% confidence.

```{r}
poverty_tested <- poverty_master |> 
  mutate(
    change   = `share_2017` - `share_2010`,
    se_2010  = `moe_2010` / 1.645,
    se_2017  = `moe_2017` / 1.645,
    se_diff  = sqrt(se_2010^2 + se_2017^2),
    moe_diff = se_diff * 1.645,
    z        = change / se_diff,
    significant = abs(z) > 1.645    #using z-score for 90% confidence (2 tails)
  )

#Removing NA values
tested_NA <- poverty_tested |> 
  filter(change == "NaN")

tested_NA

```
[_Above: Detected NA values from dataframe_]{style="color: green;"}

8 tracts displayed NA values for poverty shares for one/both surveys. To figure out what caused this, I input their GEOID into the [census website](https://data.census.gov/map/1400000US42101005000%2C42101980300%2C42101980400%2C42101980500%2C42101980600%2C42101980700%2C42101980800%2C42101980900) to check their location. We can see that these tracts are located in non-residential areas. For example,

* Tract 9804 covers the airport area
* Tract 9806 covers the parking lot in Stadium
* Tract 9808 covers Overbrooke Park

Therefore we will remove them from the dataset because it is irrelevant to poverty analysis of residents in Philadelphia.

**For the remaining rows, we determine the number of significant changes:**


```{r}

poverty_tested <- poverty_tested |> 
    filter(change != "NaN")

poverty_tested |> 
  count(significant)

```
[_Above: Count of tracts with a statistically significant change in poverty rate from 2010 to 2017_]{style="color: green;"}

Only 63 out of 376 total tracts show a statistically detectable change across the 2 ACS surveys, which could means that the article's claims are not very valid.

### 2e. Visualising the change in poverty rates: A bar chart of the 15 most reliable tracts

``` {r}
poverty_tested |> 
  arrange(desc(moe_diff)) |> 
  slice_tail(n = 15) |>                                        # Of the 15 values with LOWEST moe
  mutate(NAME = str_remove(NAME, "Census Tract "),
         change = change * 100,
         moe_diff = moe_diff * 100) |> 
  ggplot(aes(x = reorder(NAME, change), y = change)) +
  geom_col(fill = "forestgreen") +
  geom_errorbar(aes(ymin = change - moe_diff, ymax = change + moe_diff), width = 0.3) +
  coord_flip() +
  theme_minimal() +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Change in poverty rate (%) from 2006-2010 to 2013-2017",
    subtitle = "ACS 2023 5-year estimates, for the top 15 most statistically reliable tracts
    (based on lowest MoE)",
    x = "Census Tract number",
    y = "Change in poverty rate (%)"
  )

```
[_Above: Bar chart (with error bars) of poverty change in top 15 most reliable tracts_]{style="color: green;"}

The margins of error (displayed in the error bars) is very large because it is the sum of the margins of error for both surveys.


***

# Part 3 — Adjudicate
Hence we will compare the article's city-wide and neighbourhood-specific claims against the statistical significance of the change we identified.

### 3a. Supported

::: {.callout-note icon=false}
## Claim 1
"Philadelphia has a 26% poverty rate from 2013-2017."
:::


``` {r}

# Calculate overall mean, moe, cv
poverty_tested |> 
  summarise(
    number_of_tracts   = n(),
    mean_share = mean(share_2017),
    mean_moe   = (moe_sum(moe = moe_2017, estimate = share_2017) / n()) ,
    mean_cv = (mean_moe / 1.645)/mean_share
  )
```
The mean poverty rate across all tracts is indeed 26%, with a low coefficient of variation (cv) of 1.74%, indicating high reliability.

::: {.callout-note icon=false}
## Claim 2
"Fairhill in North Philadelphia registered a stunning 61 percent poverty rate."
:::

Using the Census Geocoder, the approximate tract of Fairhill is Census Tract 176.02.

``` {r}

# Calculate overall mean, moe, cv
poverty_tested |> 
  filter(NAME == "Census Tract 176.02") |> 
  select(GEOID, NAME, share_2017, moe_2017) |> 
  mutate(cv = (moe_2017 / 1.645)/share_2017)
```
Based on Jurjevich et al. (2018), the calculated poverty rate for Fairhill is accurate, but only somewhat reliable based on the coefficient of variation of 12%.

### 3b. Not supported

::: {.callout-note icon=false}
## Claim 3
"The neighborhood of Eastwick in Southwest Philadelphia, near Philadelphia International Airport, showed the greatest poverty-rate increase of any community, rocketing from 12 percent during 2006-10 to 26 percent in the newest survey. That means that even after the recession ended, around a decade ago, conditions in Eastwick have gotten much worse than during the depth of the downturn."
:::

Eastwick: Census Tract 61

``` {r}

# Calculate overall mean, moe, cv
poverty_tested |> 
  filter(NAME == "Census Tract 61") |> 
  select(GEOID, NAME, share_2017, share_2010, change, moe_diff, z, significant) |> 
  mutate(cv = (moe_diff / 1.645)/change)

```
Figures quoted are comparable enough, increasing from 13% during 2006-10 to 25% in the newest survey. 
However, the coefficient of variation is 76% which is very high, suggesting that there is no statistically detectable change from 2010 to 2017.

::: {.callout-note icon=false}
## Claim 4
"White poverty throughout much of the city skyrocketed, led by the Northeast neighborhood of Holmesburg, which went from a 2 percent poverty rate to 19 percent."
:::

white poverty???

Holmesburg: Census Tract 326

``` {r}

# Calculate overall mean, moe, cv
poverty_tested |> 
  filter(NAME %in% c("Census Tract 326", "Census Tract 315.01")) |> 
  select(GEOID, NAME, share_2017, share_2010, change, moe_diff, z, significant) |> 
  mutate(cv = ((moe_diff / 1.645) / change) * 100)

```

As seen, equating Holmesburg to Tract 326 yielded very different values with a poverty rate of 17% in 2010, unlike the 2% rate that the article claimed. Exploration of values in adjacent tracts showed high variability within Holmesburg - for example, Tract 315.01 has a vastly different poverty rate of 7% in 2010.

While it is true that poverty rates "skyrocketed" to 16% and 22% across the two tracts, the significance is false. The article's statement is not only statistically invalid, but also does not account for high variability within Holmesburg.

Inconclusive. Feel like need to look at all tracts?


### 3c. Cannot be determined




***
# archive



 Reformat to wide to put the years side by side



```{r}
#| code-fold: false
# this chunk stays visible
```
And to hide a chunk entirely — useful for library loading and data pulls that nobody needs to see:

```{r}
#| include: false
library(tidycensus)
library(tidyverse)
library(gt)
```

Other chunk options worth knowing:

Option	Effect
#| echo: false	Hide the code, show the output
#| message: false	Suppress package startup messages
#| warning: false	Suppress warnings
#| fig-cap: "..."	Add a figure caption
Use message: false and warning: false on your data pull chunks. tidycensus is chatty, and its progress output does not belong in a finished report.
